\chapter*{Acknowledgements}

% supervisors
The first and foremost acknowledgement I would like to make is in thanking my supervisors - \textit{Dave Lewis} and \textit{Declan O'Sullivan}. This work has been made possible by their vision, encouragement, motivation, and guidance. \\ [10mm]

% friends, parents
I would also like to express my thanks to my parents - \textit{pappa} and \textit{mummy} - who always have and continue to provide support and motivation since I came to being.
Thank you \textit{Annalina Caputo} for your understanding, strength, and friendship in this journey.
% and fuck you for it too! :)
Thank you \textit{Gauri Noolkar} for asking me to always keep moving ahead and for pushing me through the difficult times.

The environment at work and away was a joy thanks to the many people whom I met and were around. It is a pleasure to know you - \textit{Gary, Nicole, Lucy, Ademar, Brendan, Jamie, Javier, Kris, Christoph, Anirban, Ramisa, Anuj, Dominika, Sahil, Judie, Jeremy}, and many others. 

Thank you to \textit{Sabrina, Axel, Javier} for giving me the opportunity to visit Vienna and work there. Thanks to \textit{Rob, Christoph}, and \textit{Eoin} for providing guidance and assistance, and to \textit{Mark} for pushing me to apply myself in the real world. \\[10mm]

% funding
This work is supported by the ADAPT Centre for Digital Content Technology which is funded under the SFI Research Centres Programme (Grant 13/RC/2106) and is co-funded under the European Regional Development Fund.
